A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-04 paper/paper.tex
prose for the profunctor connectives

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorintroatXwithYandZ}[3]{(#2,#1,#3)}
\newcommand{\tensorelimXtoYwithkontZ}[3]{\textrm{ind}_{\to}(#1,#2,#3)}
\newcommand{\tensorelimfromWXtoYwithkontZ}[4]{\textrm{ind}_{\odot}(#1,#2,#3,#4)}
\newcommand{\tensorelimVatWtoXtoYwithkontZ}[5]{\textrm{ind}_{\odot}(#4,#2,#1,#3,#5)}
\newcommand{\tensorelimWfromXtoYkontZ}[4]{\textrm{ind}_{\odot}(#4;#1,#2,#3)}

@@

Next, we cover the connectives for the sets, i.e., constructions on
profunctors in Figure~\ref{fig:set-connectives}.

@@

    \inferrule
    {\Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q}

    \inferrule
    {\Gamma \pipe \alpha:\cat C; d^-\Xi\vdash R \\\\ \Gamma \pipe \alpha : \cat C; d^+\Xi\vdash P}
    {\Gamma \pipe \Xi \vdash \homlallXYtoZ \alpha R P}

  \inferrule
  {\Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R \\\\ \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P}
  {\Gamma \pipe \Xi \vdash \homrallXYtoZ \alpha R P}\\

@@

  \inferrule
  {
   \Gamma\pipe \alpha:\cat C, x:P, \beta:\cat D, y:Q, \gamma:\cat E \vdash t : R\\\\
   \Gamma \pipe\Phi \vdash s : \tensorexistsXwithYandZ {\beta} {P[a/\alpha]}{Q[c/\gamma]}}
  {\Gamma \vdash \tensorelimWfromXtoYkontZ a s b {\alpha.x.\beta.y.\gamma. t} : R[a/\alpha;c/\gamma]}

@@

Next, we develop the basics of the model theory for \ohol{}.
%
First, we demonstrate that a hyperdoctrine of virtual double
categories with split restrictions provides a sound and complete
notion of categorical model.
%
Then we instantiate this general notion of model to show that the
\ohol{} can be interpreted in ordinary category theory as well as
enriched and internal category theory, and more trivially, in terms of
sets and relations or preordered sets and ordered relations.

>>>>

\newcommand{\odotexists}[1]{\mathop{\overset{\exists #1}\odot}}
\newcommand{\tensorexistsXwithYandZ}[3]{#2 \odotexists{#1} #3}
\newcommand{\tensorintroofXandY}[2]{\textrm{pair}_\odot(#1,#2)}
\newcommand{\tensorintroatXwithYandZ}[3]{(#2,#1,#3)}
\newcommand{\tensorelimXtoYwithkontZ}[3]{\textrm{ind}_{\to}(#1,#2,#3)}
\newcommand{\tensorelimfromWXtoYwithkontZ}[4]{\textrm{ind}_{\odot}(#1,#2,#3,#4)}
\newcommand{\tensorelimVatWtoXtoYwithkontZ}[5]{\textrm{ind}_{\odot}(#4,#2,#1,#3,#5)}
\newcommand{\tensorelimWfromXtoYkontZ}[4]{\textrm{ind}_{\odot}(#4;#1,#2,#3)}
\newcommand{\tensorelimWkontZ}[2]{\textrm{ind}_{\odot}(#2;#1)}

@@

Next, we cover the connectives for the sets, i.e., constructions on
profunctors in Figure~\ref{fig:set-connectives}.
%
The top of the figure includes the formation rules for tensor products
$\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
\alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$.
%
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$.
%
Its introduction and elimination rules are like those of an
\emph{ordered} variant of equality: the introduction rule is the
identity morphism (reflexivity) and the elimination rule is an
induction principle that is a variant of the J rule of equality
elimination: we can use a term of $s : \punitinXfromYtoZ {\cat C} a b$
by specifying the behavior when $s$ is of the form $\punitrefl \alpha$
in the form of a continuation $\alpha. t$. Again we elide all
$\beta\eta$ rules, which are standard. Next, the tensor product of
sets is a kind of combined existential quantifier and monoidal
product, which we combine into a single notation
$\tensorexistsXwithYandZ \beta P Q$, where $\beta$ is the covariant
variable of $P$ and the contravariant variable of $Q$. Then the
covariant variable of the tensor product is the covariant variable of
$Q$ and the contravariant variable similarly comes from $P$.
%
In ordinary category theory, this is the \emph{composition} of
profunctors, and is defined by a coend of a product. We require that
the variable $\beta$ quantifies over a small category $\cat D$, as in
general this composite doesn't exist for large categories.
%
The introduction and elimination are like those for a combined tensor
product and existential type: the introduction rule is a pair of
terms, with an appropriate instantiation of $\beta$, and the
elimination rule says to use a term of a tensor product, it is
sufficient to specify the behavior on two elements typed with an
arbitrary middle object $\beta$.
%
Note that in the unit elimination, the continuation is typed with a
context consisting of only a single object $\alpha$, in contrast with
normal rules for a unit type in natural deduction for ordered linear
logic such as \cite{polakawpfenning}, which would split the context
$\Phi$ into $\Phi_l \jnctx \Phi_m \jnctx \Phi_r$ where $\Phi_m$ is
used by the scrutinee and $\Phi_l \jnctx \Phi_r$ is used by the
continuation. However, unlike in ordered linear logic, the $\Phi_l
\jnctx \Phi_r$ is not, necessarily well formed because there is no
reason in general that $\Phi_m$ has contravariant and covariant
dependence on the same category.
%
This is analogous to a similar situation in dependent type theory for
identity types: TODO: Dan please help with this explanation!!

Next, we introduce the contravariant ($\homlallXYtoZ \alpha R P$) and
covariant ($\homrallXYtoZ \alpha R P$) homs of sets, which are
different because we are in an orderd logic.
%
These are a kind of universally quantified
function type, where the universally quantified variable must occur
with the same variance in domain and codomain. In the contravariant
case, it occurs as the contravariant variable in both, and vice-versa
for the covariant case.
%
To highlight this, the notation for the contravariant dependence puts
the quantified variable on the \emph{left} of the triangle, as
contravariant variables occur to the left of the covariant variable,
and similarly the covariant hom has the quantified variable on the
right.
%
Then the
covariant variable of the contravariant hom set is the covariant
variable of the codomain and, and the contravariant variable of the
hom set is the \emph{covariant} variable of the domain, as the two
contravariances cancel. The covariant hom is dual.
%
Semantically, in ordinary category theory these are known as the
\emph{hom} of profunctors and are constructed using a combination of
an end and a function set.
%
The two connectives have similar introduction and elimination rules in
the form of $\lambda$ terms abstracting over both the object of the
category and the element of the set, and appropriate application
forms.
%
To keep with our invariant that the variable occurrences occur left to
right in the term syntax in a manner matching the context, we write
the covariant application in the usual order $\homrappXtoYatZ s t a$
where the function is on the left and the argument is on the right,
and the contravariant application in the flipped order.

Finally, we have the cartesian unit and product sets, which are
analogous to the normal unit and product of types. The most notable
point to emphasize is that in the formation rule for the product, the
two subformulae should have the same covariant and contravariant
dependence. So as with linear logic, the cartesian unit and product
complicate the syntactic formulation of linearity: the unit
constructor ``uses'' all variables and the product ``uses'' all
variables in both cases.

@@

    \inferrule
    {\cat D \isSmall\\\\
      \Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q}

    \inferrule
    {d^-\Xi \isSmall\\\\
      \Gamma \pipe \alpha:\cat C; d^-\Xi\vdash R \\\\ \Gamma \pipe \alpha : \cat C; d^+\Xi\vdash P}
    {\Gamma \pipe \Xi \vdash \homlallXYtoZ {\alpha: \cat C} R P}

  \inferrule
  {d^+\Xi \isSmall\\\\ \Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R \\\\ \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P}
  {\Gamma \pipe \Xi \vdash \homrallXYtoZ {\alpha:\cat C} R P}\\

@@

  \inferrule
  {
   \Gamma\pipe \Phi_m \jnctx x:P, \beta:\cat D, y:Q \jnctx \Phi_r \vdash t : R\\\\
   \Gamma \pipe\Phi_m \vdash s : \tensorexistsXwithYandZ {\beta} P Q}
  {\Gamma\pipe\Phi_l\jnctx \Phi_m \jnctx \Phi_r \vdash \tensorelimWkontZ s {x.\beta.y. t} : R}

@@

Next, we develop the basics of the model theory for \ohol{}.
%
First, we demonstrate that a hyperdoctrine of virtual equipments
categories with split restrictions provides a sound and complete
notion of categorical model.
%
Then we instantiate this general notion of model to show that the
\ohol{} can be interpreted in ordinary category theory as well as
enriched and internal category theory, and more trivially, in terms of
sets and relations or preordered sets and ordered relations.
